---
title: "Making QQ plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{qq-plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1)
```

We often make QQ plots in the Katsevich Lab, and we've found the `qqplotr` package to be somewhat deficient for this purpose. In particular, this package breaks when multiple QQ plots are displayed on the same axes, or when axis transformations are used. The `katlabutils` package therefore contains functions to help us make QQ plots with the aforementioned functionalities.
```{r setup, message=FALSE}
library(katlabutils)
library(ggplot2)
library(dplyr)
library(tibble)
```

Let's start with a basic example of a vanilla QQ plot for uniform random variables.
```{r}
data <- tibble(pvalue = runif(100))
data |>
  ggplot(aes(y = pvalue)) + # variable to be plotted in y aesthetic
  stat_qq_points() +        # add the points to the QQ plot
  stat_qq_band() +          # add the band to the QQ plot
  geom_abline() +           # add the 45 degree line
  labs(
    x = "Expected quantile",
    y = "Observed quantile"
  ) +
  theme_bw()
```

If we want to look at the tail of the distribution, it's useful to log-transform both axes. We can do this with the help of the `revlog_trans()` function:
```{r}
data |>
  ggplot(aes(y = pvalue)) +
  stat_qq_points() +
  stat_qq_band() +
  geom_abline() +
  labs(
    x = "Expected quantile",
    y = "Observed quantile"
  ) +
  scale_x_continuous(trans = revlog_trans(base = 10)) +
  scale_y_continuous(trans = revlog_trans(base = 10)) +
  theme_bw()
```

We can create QQ plots for arbitrary [distributions from the stats package](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Distributions.html). We just need to specify the distribution to the `distribution` argument to both `stat_qq_points` and `stat_qq_band`. If the `stats` package has density function `dxxxx`, then we pass `distribution = "xxxx"` to `stat_qq_points` and `stat_qq_band`. Below is an example of a normal QQ plot:
```{r}
data <- tibble(zvalue = rnorm(100))
data |>
  ggplot(aes(y = zvalue)) +
  stat_qq_points(distribution = "norm") +
  stat_qq_band(distribution = "norm") +
  geom_abline() +
  labs(
    x = "Expected quantile",
    y = "Observed quantile"
  ) +
  theme_bw()
```

The `stat_qq_points` and `stat_qq_band` layers can be composed arbitrarily with `ggplot2`'s functionalities. For example, we can display multiple QQ plots on the same axes using color:
```{r, fig.width=4}
data <- tibble(
  pvalue = c(runif(100), sqrt(runif(100))),
  method = c(rep("A", 100), rep("B", 100))
)
data |>
  ggplot(aes(y = pvalue, colour = method)) +
  stat_qq_points() +
  stat_qq_band() +
  geom_abline() +
  labs(
    x = "Expected quantile",
    y = "Observed quantile"
  ) +
  scale_x_continuous(trans = revlog_trans(base = 10)) +
  scale_y_continuous(trans = revlog_trans(base = 10)) +
  theme_bw()
```

We can also using `ggplot2`'s faceting functionality:

```{r, fig.width=6}
data <- rbind(
  tibble(
    pvalue = c(runif(100), sqrt(runif(100))),
    method = c(rep("A", 100), rep("B", 100)),
    dataset = c("Dataset 1")
  ),
  tibble(
    pvalue = c(runif(100), runif(100)),
    method = c(rep("A", 100), rep("B", 100)),
    dataset = c("Dataset 2")
  )
)
data |>
  ggplot(aes(y = pvalue, colour = method)) +
  stat_qq_points() +
  stat_qq_band() +
  geom_abline() +
  scale_x_continuous(trans = revlog_trans(base = 10)) +
  scale_y_continuous(trans = revlog_trans(base = 10)) +
  facet_wrap(~dataset) +
  labs(
    x = "Expected quantile",
    y = "Observed quantile"
  ) +
  theme_bw()
```
